---
title: "Data Importing and “Tidy” Data"
author: "El Mex"
output: 
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Basic functions for data tidying

```{r}
# import the libraries to be used
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(nycflights13)
library(fivethirtyeight)
```
## Using `readr`

```{r}
# import the ".csv" file by using "read_csv()" function
library(readr)
dem_score <- read_csv("dem_score.csv")

head(dem_score)

```

## “Tidy” data
```{r}
# explore the data frame "drinks" from the "fivethirtyeight" package
glimpse(drinks)
head(drinks)
```

**Apply these actions:**

1. `filter()` the drinks data frame to only consider 4 countries: USA, China, Italy, and Saudi Arabia.

1. `select()` all columns except `total_litres_of_pure_alcohol` by using the - sign

1. `rename()` the variables `beer_servings`, `spirit_servings`, and `wine_servings` to `beer`, `spirit`, and `wine`, respectively. 

1. Save all in a variable called `drinks_smaller`

```{r}
drinks_smaller <- drinks %>% 
  filter(country %in% c("USA", "China", "Italy", "Saudi Arabia")) %>% 
  select(-total_litres_of_pure_alcohol) %>% 
  rename(beer = beer_servings, spirit = spirit_servings, wine = wine_servings)

drinks_smaller
```
## Converting to "tidy" data

**Most original data frames are in wide (non-“tidy”) format as it is above.  Use `pivot_longer()` function in the `tidyr` package to convert it to "tidy" format**

**Reminder: **

1. `names_to` is the name of the variable in the new “tidy”/long data frame that will contain the *column names* of the original data

1. `values_to` is the name of the variable in the new “tidy” data frame that will contain the *values* of the original data

1. `cols` is the columns in the data frame you either want to or don’t want to “tidy”

```{r}
# We don't want to tidy it to the "country" column so we put "-" sign to country (-country)
drinks_smaller_tidy <- drinks_smaller %>% 
  pivot_longer(names_to = "type",
               values_to = "servings",
               cols = -country)

drinks_smaller_tidy
```

**As the variable names of `type` & `servings` don’t appear in `drinks_smaller`, we use quotation marks around it, on the other hand, `country` appear in `drinks_smaller` so no need for quotation marks**

```{r eval=FALSE}
# another way to get the same results
drinks_smaller %>% 
  pivot_longer(names_to = "type",
               values_to = "servings",
               cols = c(beer, spirit, wine))

# since these three columns appear one after another in "drinks_smaller", we could also do this:
drinks_smaller %>% 
  pivot_longer(names_to = "type", 
               values_to = "servings", 
               cols = beer:wine)
```

**Create the side-by-side barplot:**

1. The categorical variable `country` with four levels (China, Italy, Saudi Arabia, USA) would have to be mapped to the `x`-position of the bars

1. The numerical variable `servings` would have to be mapped to the `y`-position of the bars (the height of the bars)

1. The categorical variable `type` with three levels (beer, spirit, wine) would have to be mapped to the `fill` color of the bars

```{r}
ggplot(drinks_smaller_tidy, aes(x = country, y = servings, fill = type)) +
  geom_col(position = "dodge")
```
**Let's do another exercise using the `airline_safety` data frame in the `fivethirtyeight` package**

```{r}
# explore it
glimpse(airline_safety)
```

```{r}
airline_safety_smaller <- airline_safety %>% 
  select(airline, starts_with("fatalities"))

airline_safety_smaller
```
**How would you convert this data frame to be in “tidy” format, in particular so that it has a variable `fatalities_years` indicating the incident year and a variable `count` of the fatality counts?**

```{r}
airline_tidy <- airline_safety_smaller %>% 
  pivot_longer(names_to = "fatalities_years",
               values_to = "count",
               cols = -airline)

airline_tidy
```

## Case study: Democracy in Guatemala

**Let’s use the `dem_score` data frame we imported, but focus on only data corresponding to Guatemala**

```{r}
guat_dem <- dem_score %>% 
  filter(country == "Guatemala")

guat_dem
  
```


**Our data isn't tidy. Create a data frame with three columns: `country`, `year`, and `democracy_score`**


```{r}
 guat_dem_tidy <- guat_dem %>% 
  pivot_longer(names_to = "year",
               values_to = "democracy_score",
               cols = -country,
               names_transform = list(year = as.integer))

guat_dem_tidy
```

**The original column names ('1952', '1957'...) were made of a list of characters, so we had to convert the new variable `year` as integers**

**Create a time-series plot**

```{r}
ggplot(guat_dem_tidy, aes(x = year, y = democracy_score)) +
  geom_line() +
  labs(x = "Year", y = "Democracy Score")
```
















